Haskell in Computational Biology

### 2013-05-29 jfp.tex
tighten the Senington story

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 2,500 lines of~Haskell, about 500 of which are devoted
to tests, QuickCheck properties, and generators.
Neither Matt nor SMURF includes test code.
MRFy~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
\mrfy's initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three tools were written
 in~haste.
We~have experience modifying all three tools.

@@

We~built MRFy quickly, and we expected that
higher-order functions would make MRFy easy to extend.
Each new addition to MRFy's stochastic search took at most a day
to implement.
However, the next extension we wished to make
was based on
Richard Senington's model of \emph{local
search} \cite{senington:combinators-local}, 
which does not fit our abstraction of stochastic search.
We~wound up writing a new program that combines parts
of MRFy with Senington's  \texttt{local-search} package.
This~task took only three weeks, and the technical results were terrific.
Local search outperforms random hill-climbing, simulated
annealing, and the genetic algorithm.
Even though we're not entirely happy with the
state of the resulting code, we~count the story as a success.

>>>>

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 2,500 lines of~Haskell, about 500 of which are devoted
to tests, QuickCheck properties, and generators.
Neither Matt nor SMURF includes test code.
MRFy~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
\mrfy's initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three tools were written
 in~haste.
And all three have later been modified.

@@

We~built MRFy quickly, and we expected that
higher-order functions would make MRFy easy to extend.
Each new addition to MRFy's stochastic search took at most a day
to implement.
However, the next extension we wished to make
was based on
Richard Senington's model of \emph{local
search} \cite{senington:combinators-local}, 
which does not fit our abstraction of stochastic search.
We~wound up writing a new program that adapts parts
of MRFy to work with Senington's  \texttt{local-search} package.
We're~disappointed that such similar algorithms have entirely
different interfaces,
but the adaptation took only three weeks, and the
technical results were terrific. 
Local search outperforms random hill-climbing, simulated
annealing, and the genetic algorithm.
We~therefore count the story as a success.
